---
title: "R 4 Data Science"
author: "Andi Shehu"
date: "11/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = FALSE, 
                      warning = FALSE)
```


#  Chapter 5

## 5.2.4 Exercises

1. Find all flights that

```{r}
library(tidyverse)
library(nycflights13)
# 1.Had an arrival delay of two or more hours
names(flights)

flights %>% 
  filter(arr_time > 120)

#2. Flew to Houston (IAH or HOU)

flights %>% 
  filter(dest %in% c("IAH", "HOU"))

#3. Were operated by United, American, or Delta

unique(flights$carrier)
flights %>% 
  filter(carrier %in% c("AU", "AA", "DL"))

#4. Departed in summer (July, August, and September)

unique(flights$month)

flights %>% 
  filter(month %in% 7:9)

```


5. Arrived more than two hours late, but didn’t leave late

Ans. 

```{r}
flights %>% 
  filter(arr_delay > 120 & dep_delay <=0)

```

1.6 Were delayed by at least an hour, but made up over 30 minutes in flight

Ans. dep_del >= 60 AND 
made up over 30 minutes in flight== basically arrived at least 30 min early


```{r}
flights %>% 
  filter(dep_delay >= 60 & arr_delay < 30)

```


1.7 Departed between midnight and 6am (inclusive)
```{r}
flights %>% 
  filter(dep_time %in% c(2400, 1:600))

```


2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?

Ans. No 
```{r}
flights %>% 
  filter(between(dep_time,1, 600))
```


3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
#a
flights %>% 
  filter(is.na(dep_time)) %>% 
  dim()

#What other variables are missing?
flights %>% 
  filter(is.na(dep_time))

```

dep_delay and arr_time have the same missing values 
They probably represent flights that were cancelled. 




4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)



# 5.3.1 Exercises

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}

flights %>% 
  arrange(desc(is.na(dep_delay)))

```


2. Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
#most delayed
flights %>% 
  arrange(desc(dep_delay)) %>% 
  slice(1:5)

#left earliest 

flights %>% 
  arrange(dep_time)

```


3. Sort flights to find the fastest (highest speed) flights.

Speed is actually not a variable but we can create one 
speed = distance/time 


```{r}
names(flights)

flights %>% 
  mutate(speed = distance/air_time, na.rm = TRUE) %>% 
  arrange(desc(speed))

```

Can also be answered without creating a new variable but this way I can check on speed 

4. Which flights travelled the farthest? Which travelled the shortest?

```{r}

#furthest == means most distance 
flights %>% 
  arrange(desc(distance))

#shortest flights 

flights %>% 
  arrange(distance) 

```

17 miles for shortest - that should just be a drive. 


## 5.4.1 Exercises

1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}
flights %>% 
  select(dep_time, dep_delay, arr_time, arr_delay)
```


2. What happens if you include the name of a variable multiple times in a select() call?

```{r}
flights %>% 
  select(dep_time, dep_time)
```

Nothing - gets ignored

3. What does the any_of() function do? Why might it be helpful in conjunction with this vector?


```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

flights %>% 
  select(one_of(vars))

```


4. Does the result of running the following code surprise you?
How do the select helpers deal with case by default? 

```{r}
select(flights, contains("TIME"))
```


Ans.
Yes. Apparently the function `contains()` picks up both lower case and upper case 

How can you change that default?

```{r}

select(flights, contains("TIME", ignore.case = FALSE))

```

Ans. set `ignore.case` to FALSE 

## 5.5.2 Exercises

1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

```{r}
#use transmute to check 
flights %>% 
  transmute(hour = dep_time %/%100,
            min = dep_time%%100,
            dep_time_min = hour*60+min)

```


2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?


```{r}
flights %>% 
  mutate(air_time_calc = arr_time - dep_time) %>% 
  select(air_time, air_time_calc)
```


They should be equal - we see that the calculated air time is actually higher than the given air_time. Convert both arr_time and dep_time in minutes 


```{r}

flights %>% 
  mutate(arr_time_min = (arr_time%/%100)*60 + arr_time%%100, 
         dep_time_min = (dep_time%/%100)*60 + dep_time%%100,
         air_time_calc = arr_time_min - dep_time_min) %>% 
  select(air_time_calc, air_time)

```


They are still not equal...

3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

```{r}
flights %>% 
  select(dep_time, sched_dep_time, dep_delay)

```

should be 

`dep_time` - `sched_dep_time` = `dep_delay`


```{r}
flights %>% 
  mutate(dep_delay_calc = dep_time - sched_dep_time) %>% 
  select(dep_delay_calc, dep_delay)

```



4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().


```{r}

flights %>% 
  transmute(delay_time = dep_delay+arr_delay) %>% 
  mutate(ranked_rows = row_number(delay_time)) %>% 
  arrange(desc(ranked_rows)) %>% 
  slice(1:10)
  

```


5. What does 1:3 + 1:10 return? Why?

```{r}
1:3+1:10

```


6. What trigonometric functions does R provide?

basic trig such as sin, cos, tan


```{r}
sin(45)
```


## 5.6.7 Exercises

1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:

  * A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.

  *A flight is always 10 minutes late.

```{r}

#label flights which are ten min late
flights %>% 
  mutate(ten_min_late = ifelse(arr_delay >10, "Yes", "No")) %>% 
  select(arr_delay, ten_min_late)

```



  *A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.



  *99% of the time a flight is on time. 1% of the time it’s 2 hours late.


Which is more important: arrival delay or departure delay?

Ans. Arrival delay. Because the flights can depart late, and make up time on flight. 


2. Come up with another approach that will give you the same output as not_cancelled %>% count(dest) and not_cancelled %>% count(tailnum, wt = distance) (without using count()).




3. Our definition of cancelled flights (is.na(dep_delay) | is.na(arr_delay) ) is slightly suboptimal. Why? Which is the most important column?



4. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?


5. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights %>% group_by(carrier, dest) %>% summarise(n()))



6. What does the sort argument to count() do. When might you use it?


















